---
title: "simGH: simulating times to event from a general hazard structure"
author: '[F. Javier Rubio](https://sites.google.com/site/fjavierrubio67/)'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
bibliography: references.bib
---

# The `simGH` command.
The simGH command from the `HazReg` R package allows one to simulate times to event from the following models:

- General Hazard (GH) model [@chen:2001] [@rubio:2019].

- Accelerated Failure Time (AFT) model [@kalbfleisch:2011].

- Proportional Hazards (PH) model [@cox:1972].

- Accelerated Hazards (AH) model [@chen:2000].


A description of these hazard models is presented below as well as the available baseline hazards.

## General Hazard model
The GH model is formulated in terms of the hazard structure
\[
h(t; \alpha, \beta, \theta, {\bf x}) = h_0\left(t  \exp\{\tilde{\bf x}^{\top}\alpha\}; \theta\right) \exp\{{\bf x}^{\top}\beta\}.
\]
where ${\bf x}\in{\mathbb R}^p$ are the covariates that affect the hazard level; $\tilde{\bf x} \in {\mathbb R}^q$ are the covariates the affect the time level (typically $\tilde{\bf x} \subset {\bf x}$); $\alpha \in {\mathbb R}^q$ and $\beta \in {\mathbb R}^p$ are the regression coefficients; and $\theta \in \Theta$ is the vector of parameters of the baseline hazard $h_0(\cdot)$.

This hazard structure leads to an identifiable model as long as the baseline hazard is not a hazard associated to a member of the Weibull family of distributions [@chen:2001]. 

## Accelerated Failure Time (AFT) model
The AFT model is formulated in terms of the hazard structure
\[
h(t; \beta, \theta, {\bf x}) = h_0\left(t  \exp\{{\bf x}^{\top}\beta\}; \theta\right) \exp\{{\bf x}^{\top}\beta\}.
\]
where ${\bf x}\in{\mathbb R}^p$ are the available covariates; $\beta \in {\mathbb R}^p$ are the regression coefficients; and $\theta \in \Theta$ is the vector of parameters of the baseline hazard $h_0(\cdot)$.

## Proportional Hazards (PH) model
The PH model is formulated in terms of the hazard structure
\[
h(t; \beta, \theta, {\bf x}) = h_0\left(t ; \theta\right) \exp\{{\bf x}^{\top}\beta\}.
\]
where ${\bf x}\in{\mathbb R}^p$ are the available covariates; $\beta \in {\mathbb R}^p$ are the regression coefficients; and $\theta \in \Theta$ is the vector of parameters of the baseline hazard $h_0(\cdot)$.

## Accelerated Hazards (AH) model
The AH model is formulated in terms of the hazard structure
\[
h(t; \alpha, \theta, \tilde{\bf x}) = h_0\left(t \exp\{\tilde{\bf x}^{\top}\alpha\}; \theta\right) .
\]
where $\tilde{\bf x}\in{\mathbb R}^q$ are the available covariates; $\alpha \in {\mathbb R}^q$ are the regression coefficients; and $\theta \in \Theta$ is the vector of parameters of the baseline hazard $h_0(\cdot)$.

# Available baseline hazards
The current version of the `simGH` command implements the following parametric baseline hazards for the models discussed in the previous section.

- [Power Generalised Weibull](http://rpubs.com/FJRubio/PGW) (PGW) distribution.

- [Exponentiated Weibull](http://rpubs.com/FJRubio/EWD) (EW) distribution.

- [Generalised Gamma](http://rpubs.com/FJRubio/GG) (GG) distribuiton.

- [Gamma](https://en.wikipedia.org/wiki/Gamma_distribution) (G) distribution.

- [Lognormal](https://en.wikipedia.org/wiki/Log-normal_distribution) (LN) distribution.

- [Log-logistic](https://en.wikipedia.org/wiki/Log-logistic_distribution) (LL) distribution.

- [Weibull](https://en.wikipedia.org/wiki/Weibull_distribution) (W) distribution. (only for AFT, PH, and AH models)


# Illustrative example: R code
In this example, we simulate $n=1,000$ times to event from the GH, PH, AFT, and AH models with PGW baseline hazards, using the `simGH` command. We censor these samples at a fixed value, and fit the corresponding models using the R package `HazReg`.

See also: [HazReg](https://github.com/FJRubio67/HazReg)

## PGW-GH model
```{r}
rm(list=ls())

# Required packages
#library(devtools)
#install_github("FJRubio67/HazReg")
library(HazReg)

# Sample size
n <- 1000

# Simulated design matrices
set.seed(1234)
x <- as.matrix(cbind(rnorm(n), rnorm(n)))
xt <- as.matrix(rnorm(n))

#----------------------------
# PGW-GH simulation
#----------------------------

# True parameters
theta0 <- c(0.1,2,5)
betat0 <- c(0.5)
betah0 <- c(-0.5,0.75)

# censoring
cens <- 10

# Data simulation
simdat <- simGH(seed = 1234, n = n, des_h = x, des_t = xt,
      theta = theta0, beta_t = betat0, beta_h = betah0, hstr = "GH", baseline = "PGW")

# status variable
status <- (simdat < cens)

# Inducing censoring
simdat <- ifelse(simdat < cens, simdat, cens)

# Model fit
OPTPGWGH <- GHMLE(init = rep(0, 3 + ncol(x) + ncol(xt)), times = simdat, status = status,
                   hstr = "GH", dist = "PGW", des = x, des_t = xt, method = "nlminb", maxit = 10000)

MLEPGWGH <- c(exp(OPTPGWGH$OPT$par[1:3]),OPTPGWGH$OPT$par[-c(1:3)])

# True parameter values vs. MLE
cbind(c(theta0,betat0,betah0),MLEPGWGH)

```

## PGW-PH model
```{r}
rm(list=ls())

# Required packages
#library(devtools)
#install_github("FJRubio67/HazReg")
library(HazReg)

# Sample size
n <- 1000

# Simulated design matrices
set.seed(1234)
x <- as.matrix(cbind(rnorm(n), rnorm(n)))

#----------------------------
# PGW-PH simulation
#----------------------------

# True parameters
theta0 <- c(0.1,2,5)
beta0 <- c(-0.5,0.75)

# censoring
cens <- 10

# Data simulation
simdat <- simGH(seed = 1234, n = n, des = x,
                theta = theta0, beta = beta0, hstr = "PH", baseline = "PGW")

# status variable
status <- (simdat < cens)

# Inducing censoring
simdat <- ifelse(simdat < cens, simdat, cens)

# Model fit
OPTPGWPH <- GHMLE(init = rep(0, 3 + ncol(x) ), times = simdat, status = status,
                   hstr = "PH", dist = "PGW", des = x, method = "nlminb", maxit = 10000)

MLEPGWPH <- c(exp(OPTPGWPH$OPT$par[1:3]),OPTPGWPH$OPT$par[-c(1:3)])

# True parameter values vs. MLE
cbind(c(theta0,beta0),MLEPGWPH)

```

## PGW-AFT model
```{r}
rm(list=ls())

# Required packages
#library(devtools)
#install_github("FJRubio67/HazReg")
library(HazReg)

# Sample size
n <- 1000

# Simulated design matrices
set.seed(1234)
x <- as.matrix(cbind(rnorm(n), rnorm(n)))

#----------------------------
# PGW-AFT simulation
#----------------------------

# True parameters
theta0 <- c(0.1,2,5)
beta0 <- c(-0.5,0.75)

# censoring
cens <- 10

# Data simulation
simdat <- simGH(seed = 1234, n = n, des = x,
                theta = theta0, beta = beta0, hstr = "AFT", baseline = "PGW")

# status variable
status <- (simdat < cens)

# Inducing censoring
simdat <- ifelse(simdat < cens, simdat, cens)

# Model fit
OPTPGWAFT <- GHMLE(init = rep(0, 3 + ncol(x) ), times = simdat, status = status,
                   hstr = "AFT", dist = "PGW", des = x, method = "nlminb", maxit = 10000)

MLEPGWAFT <- c(exp(OPTPGWAFT$OPT$par[1:3]),OPTPGWAFT$OPT$par[-c(1:3)])

# True parameter values vs. MLE
cbind(c(theta0,beta0),MLEPGWAFT)

```

## PGW-AH model
```{r}
rm(list=ls())

# Required packages
#library(devtools)
#install_github("FJRubio67/HazReg")
library(HazReg)

# Sample size
n <- 1000

# Simulated design matrices
set.seed(1234)
x <- as.matrix(cbind(rnorm(n), rnorm(n)))

#----------------------------
# PGW-AH simulation
#----------------------------

# True parameters
theta0 <- c(0.1,2,5)
beta0 <- c(-0.5,0.75)

# censoring
cens <- 10

# Data simulation
simdat <- simGH(seed = 1234, n = n, des = x,
                theta = theta0, beta = beta0, hstr = "AH", baseline = "PGW")

# status variable
status <- (simdat < cens)

# Inducing censoring
simdat <- ifelse(simdat < cens, simdat, cens)

# Model fit
OPTPGWAH <- GHMLE(init = rep(0, 3 + ncol(x) ), times = simdat, status = status,
                   hstr = "AH", dist = "PGW", des_t = x, method = "nlminb", maxit = 10000)

MLEPGWAH <- c(exp(OPTPGWAH$OPT$par[1:3]),OPTPGWAH$OPT$par[-c(1:3)])

# True parameter values vs. MLE
cbind(c(theta0,beta0),MLEPGWAH)

```